### 2013-04-28 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding to the mc background chapter

@@

In terms of particle transport, this adjoint method is equivalent to a
traditional forward method. As a result of using the adjoint system,
we modify our probabilities and weights using the \textit{adjoint
  Neumann-Ulam decomposition} of $\ve{H}$:
\begin{equation}
  \ve{H}^{T} = \ve{P} \circ \ve{W}\:,
  \label{eq:adjoint_neumann_ulam}
\end{equation}
where now we are forming the decomposition with respect to the
transpose of $\ve{H}$. We then follow the same procedure as the direct
method for forming the probability and weight matrices in the
decomposition. Using the adjoint form, probabilities should instead be
column-scaled:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
  \label{eq:adjoint_probability}
\end{equation}
such that we expect to select a new state $j$ from the current state
in the random walk $j$ by sampling column-wise (or row-wise if an
adjoint probability matrix is formed). Per
Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
defined as:
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint_weight}
\end{equation}
Using the decomposition we can then define an expectation value for
the adjoint method. Given Eq~(\ref{eq:direct_permutation_weight}) as
the weight generated for a particular random walk permutation as in
Eq~(\ref{eq:mc_walk_permutation}) and our result from
Eq~(\ref{eq:inner_product_constraint}) generated by applying the
adjoint constraints, the contribution to the solution in state $i$
from a particular random walk permutation of $k$ events is then the
\textit{collision estimator}:
\begin{equation}
  X_{j}(\nu) = \sum_{m=0}^k W_{m} \delta_{i_m,j}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in
the current state the random walk starting weight will be $W_0 =
b_{i_0}$. Note here that the estimator in
Eq~(\ref{eq:adjoint_permutation_contribution}) does not have a
dependency on the source state as in
Eq~(\ref{eq:adjoint_permutation_contribution}), providing a remedy for
the situation in the direct method where we must start a random walk
in each source state for every permutation if we want to compute a
solution estimate for that state. In the adjoint method, we instead
tally in all states and those of lesser importance will not be visited
as frequently by the random walk. Finally, the expectation value using
all permutations is:
\begin{equation}
  E\{X_j\} = \sum_{\nu} P_{\nu} X_{j}(\nu)\:
  \label{eq:adjoint_expectation_value}
\end{equation}
which, if expanded in the same way as the direct method, directly
recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}
    \delta_{i_k,j} \\ &= x_{j}\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
therefore also providing an unbiased Monte Carlo estimate of the
solution. It should be noted here that
Eq~(\ref{eq:adjoint_expectation_expansion}) only computes a single
component of our desired solution vector when really what we desire is
the entire solution vector. In an adjoint Monte Carlo simulation using
this estimator, the $w_{i,j}$ elements that are added into the tally
for each state are only selected if the random walk currently resides
in that state. Much like a mesh tally in a particle transport
simulation, we have $N$ simultaneous tallies for $\ve{A} \in
\mathbb{R}^{N \times N}$ that will yield the entire solution
vector. Based on their current state in the system, the random walks
will contribute tally $j$ in this group.

@@

\subsection{Expected Value Estimator}
\label{subsec:expected_value_estimator}
In addition to the collision estimator, an additional estimator is
available is available due to the work of Okten
\citep{okten_solving_2005} that uses the method of expected values as
a means to improve the Monte Carlo estimate. As outlined by Spanier
and Gelbard \citep{spanier_monte_1969}, the method of expected values
is a deterministic averaging of events that may potentially occur in
the Monte Carlo random walk sequence. Okten applied this principle
directly to discrete Monte Carlo by forming the \textit{expected value
  estimator} for a random walk of $k$ events:
\begin{equation}
  X_{j}(\nu) = b_j + \sum_{m=0}^k W_m h_{j,i_m}\,
  \label{eq:expected_value_estimator}
\end{equation}
where now the contribution of the iteration matrix is
deterministically averaged at step $m$ over all potentially states $j$
that may be reached from the current state $i_m$. Via Okten, the
estimator can be shown to be unbiased through a comparison to
the expected value estimator. We can first rewrite the summation in
Eq~(\ref{eq:expected_value_estimator}):
\begin{equation}
  X_{j}(\nu) = b_j + \sum_{m=0}^k \sum_{i=1}^N W_m
  \delta_{i_m,i} h_{ji}\,
  \label{eq:unbiased_eval_1}
\end{equation}
where $N$ is the number of states in the system. Immediately, we see
the collision estimator as defined by
Eq~(\ref{eq:adjoint_permutation_contribution}) and can therefore write
the expectation value as:
\begin{equation}
  E\{X_{j}\} = b_j + \sum_{i=1}^N E\{X_{i}\} h_{ji}\,
  \label{eq:unbiased_eval_2}
\end{equation}
which is equivalently is the $j^{th}$ component of
Eq~(\ref{eq:richardson_split}):
\begin{equation}
  E\{X_{j}\} = b_j + \sum_{i=1}^N x_{i} h_{ji}\,
  \label{eq:unbiased_eval_2}
\end{equation}
and is therefore and unbiased estimate. Compared to the collision
estimator, provides additional information at every step of the random
walk, yielding potentially better statistics with the same amount of
transport work. Conveniently, even if no Monte Carlo histories are
computed, the estimator still deterministically computes the first
term of the Neumann Series, $\ve{H}^0\ve{b}$, whereas the collision
estimator will provide no information.

@@

%%---------------------------------------------------------------------------%%
\section{Estimator Variance}
\label{subsec:estimator_variance}

\subsection{Forward Estimator Variance}
\label{subsec:forward_variance}
We can compute the variance of the forward estimator through traditional
methods by defining the variance, $\sigma_i$, for each component in
the solution:
\begin{equation}
  {\sigma_i}^2 = E\{X(i_0 = i) - (\ve{A}^{-1}\ve{b})_i\}^2 = E\{X(i_0
  = i)^2\} - x_i^2\:,
  \label{eq:direct_variance_1}
\end{equation}
where the vector exponentials are computed element-wise. Inserting
Eq~(\ref{eq:direct_expectation_value}) gives:
\begin{equation}
  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} X_{\nu}^2\Big)_i - x_i^2\:,
  \label{eq:direct_variance_2}
\end{equation}
and applying Eq~(\ref{eq:direct_permutation_contribution}):
\begin{equation}
  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2
  b_{i_m}^2\Big)_i - x_i^2\:.
  \label{eq:direct_variance_3}
\end{equation}
Finally, expanding the transition probabilities yields the variance:
\begin{equation}
  \sigma_i^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k} b_{i_m} - x_i^2\:.
  \label{eq:direct_variance_4}
\end{equation}
Using this definition for the variance, we can arrive at a more natural
reason for enforcing $\rho(\ve{H}) < 1$ for our Monte Carlo method to
converge. Per the Hadamard product, we can concatenate the summation
in Eq~(\ref{eq:direct_variance_4}):
\begin{equation}
  (\ve{P} \circ \ve{W} \circ \ve{W})^k =
  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k}\:.
  \label{eq:double_weighted_decomposition}
\end{equation}
If we assign $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$ as in
Eq~(\ref{eq:neumann_ulam_decomposition}), we then have a new
formulation for the variance:
\begin{equation}
  \sigma^2_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}g_{i,i_1}g_{i_1,i_2}\ldots g_{i_{k-1},i_k} b_{i_k}^2 -
  x_i^2\:,
\end{equation}
which contains the general Neumann series for $\ve{G}$,
\begin{equation}
  \ve{T} = \sum_{k=0}^{\infty} \ve{G}^k\:,
  \label{eq:variance_neumann_series}
\end{equation}
where $\ve{T} = (\ve{I}-\ve{G})^{-1}$. We can then insert $T$ back
into the variance formulation for a more concise definition:
\begin{equation}
  \sigma^2_i = (\ve{T}\ve{b})_i - x_i^2\:.
  \label{eq:direct_variance_5}
\end{equation}
We can relate $\ve{G}$ to $\ve{H}$ by noting that $\ve{G}$ simply
contains an additional Hadamard product with the weight matrix. The
Hadamard product has the property that:
\begin{equation}
  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:.
  \label{eq:hadamard_inequality}
\end{equation}
Using the norm property of the Hadamard product and
Eq~(\ref{eq:neumann_ulam_decomposition}), we can define the norm of
$\ve{W}$ as:
\begin{equation}
  \frac{|\ve{H}|}{|\ve{P}|} \geq |\ve{W}|\:.
  \label{eq:weight_norm}
\end{equation}
Choosing the infinity norm of the operator as defined in
Eq~(\ref{eq:matrix_infinity_norm}), the row normalized probability
matrix will yield a norm of 1 giving the following inequality for
relating $\ve{G}$ and $\ve{H}$:
\begin{equation}
  |\ve{G}| \geq |\ve{H}|^2
  \label{eq:variance_norm_inequality}
\end{equation}
Using these relations to analyze Eq~(\ref{eq:direct_variance_5}), we
see that if $\rho(\ve{G}) > 1$, then the sum in
Eq~(\ref{eq:variance_neumann_series}) will not converge and an
infinite variance will arise as the elements of $\ve{T}$ become
infinite in Eq~(\ref{eq:direct_variance_5}). We must restrict $\ve{G}$
to alleviate this and therefore restrict $\ve{H}$ due to
Eq~(\ref{eq:variance_norm_inequality}) with $\rho(\ve{H}) < 1$ so that
our expectation values for the solution may have a finite variance.

\subsection{Collision Estimator Variance}
\label{subsec:collision_variance}

\subsection{Expected Value Estimator Variance}
\label{subsec:expected_value_variance}

@@

\subsection{Preconditioning MCSA}
\label{sec:stochastic_preconditioning}
In most cases, at least a minimal amount of \textit{preconditioning}
of the linear system will be required in order to use the class of
stochastic methods described. Although these methods have no symmetry
requirements for convergence, they do require that the spectral radius
of the iteration matrix be less than one. As an example, to achieve
this for diagonally dominant matrices point Jacobi left
preconditioning can be used such that the preconditioning matrix
$\ve{M}$ is:
\begin{equation}
  \ve{M} = diag(\ve{A})\:.
  \label{eq:jacobi_preconditioner}
\end{equation}
With the application of this preconditioner we are instead solving the
following linear system:
\begin{equation}
  \ve{M}^{-1}\ve{A}\ve{x} = \ve{M}^{-1}\ve{b}\:.
  \label{eq:jacobi_precond_linear_problem}
\end{equation}
Next, we can apply MCSA to solve
Eq~(\ref{eq:jacobi_precond_linear_problem}): 
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k +
    \ve{M}^{-1}\ve{r}^k\:,\\ \ve{r}^{k+1/2} =
    \ve{b}-\ve{A}\ve{x}^{k+1/2}\:,\\ \ve{M}^{-1}\ve{A}\delta\ve{x}^{k+1/2}
    = \ve{M}^{-1}\ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} +
    \delta \ve{x}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
    \label{eq:jacobi_preconditioned_mcsa}
  \end{gather}
\end{subequations}
where the adjoint Monte Carlo solve now has a preconditioned operator
from which to build weights and probabilities for transport and a
preconditioned source vector to sample.

@@

\subsection{General Preconditioning Strategies}
\label{subsec:general_mcsa_strategies}
It is possible to use general left, right, and left/right
preconditioning with MCSA by carefully considering the underly Monte
Carlo that will be solved with the Neumann Ulam method.

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/seq_cpu.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
      an $N \times N$ square mesh).} \textit{oth the Sequential Monte
      Carlo and MCSA solvers are used with the five point stencils and
      the adjoint Monte Carlo solver. The number of random walks was
      twice the number of discrete states in the system in order to
      ensure convergence in the Sequential Monte Carlo method.}}
  \label{fig:seq_cpu_time}
\end{figure}

>>>>

\subsection{Forward Estimator Variance}
\label{subsec:forward_variance}
We can compute the variance of the forward estimator through traditional
methods by defining the variance, $\sigma_i$, for each component in
the solution:
\begin{equation}
  {\sigma_i}^2 = E\{X_i - (\ve{A}^{-1}\ve{b})_i\}^2 = E\{X_i^2\} - x_i^2\:,
  \label{eq:direct_variance_1}
\end{equation}
where the vector exponentials are computed element-wise. Inserting
Eq~(\ref{eq:direct_expectation_value}) gives:
\begin{equation}
  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} X_{\nu}^2\Big)_i - x_i^2\:,
  \label{eq:direct_variance_2}
\end{equation}
and applying Eq~(\ref{eq:direct_permutation_contribution}):
\begin{equation}
  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2
  b_{i_m}^2\Big)_i - x_i^2\:.
  \label{eq:direct_variance_3}
\end{equation}
Finally, expanding the transition probabilities yields the variance:
\begin{equation}
  \sigma_i^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k} b_{i_k}^2 - x_i^2\:.
  \label{eq:direct_variance_4}
\end{equation}
Using this definition for the variance, we can arrive at a more natural
reason for enforcing $\rho(\ve{H}) < 1$ for our Monte Carlo method to
converge. Per the Hadamard product, we can concatenate the summation
in Eq~(\ref{eq:direct_variance_4}):
\begin{equation}
  (\ve{P} \circ \ve{W} \circ \ve{W})^k =
  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k}\:.
  \label{eq:double_weighted_decomposition}
\end{equation}
If we assign $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$ as in
Eq~(\ref{eq:neumann_ulam_decomposition}), we then have a new
formulation for the variance:
\begin{equation}
  \sigma^2_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}g_{i,i_1}g_{i_1,i_2}\ldots g_{i_{k-1},i_k} b_{i_k}^2 -
  x_i^2\:,
\end{equation}
which contains the general Neumann series for $\ve{G}$,
\begin{equation}
  \ve{T} = \sum_{k=0}^{\infty} \ve{G}^k\:,
  \label{eq:variance_neumann_series}
\end{equation}
where $\ve{T} = (\ve{I}-\ve{G})^{-1}$. We can then insert $T$ back
into the variance formulation for a more concise definition:
\begin{equation}
  \sigma^2_i = (\ve{T}\ve{b})_i - x_i^2\:.
  \label{eq:direct_variance_5}
\end{equation}
We can relate $\ve{G}$ to $\ve{H}$ by noting that $\ve{G}$ simply
contains an additional Hadamard product with the weight matrix. The
Hadamard product has the property that:
\begin{equation}
  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:.
  \label{eq:hadamard_inequality}
\end{equation}
Using the norm property of the Hadamard product and
Eq~(\ref{eq:neumann_ulam_decomposition}), we can define the norm of
$\ve{W}$ as:
\begin{equation}
  \frac{|\ve{H}|}{|\ve{P}|} \geq |\ve{W}|\:.
  \label{eq:weight_norm}
\end{equation}
Choosing the infinity norm of the operator as defined in
Eq~(\ref{eq:matrix_infinity_norm}), the row normalized probability
matrix will yield a norm of 1 giving the following inequality for
relating $\ve{G}$ and $\ve{H}$:
\begin{equation}
  |\ve{G}| \geq |\ve{H}|^2
  \label{eq:variance_norm_inequality}
\end{equation}
Using these relations to analyze Eq~(\ref{eq:direct_variance_5}), we
see that if $\rho(\ve{G}) > 1$, then the sum in
Eq~(\ref{eq:variance_neumann_series}) will not converge and an
infinite variance will arise as the elements of $\ve{T}$ become
infinite in Eq~(\ref{eq:direct_variance_5}). We must restrict $\ve{G}$
to alleviate this and therefore restrict $\ve{H}$ due to
Eq~(\ref{eq:variance_norm_inequality}) with $\rho(\ve{H}) < 1$ so that
our expectation values for the solution may have a finite variance.

@@

In terms of radiation transport, this adjoint method is equivalent to
a traditional forward method where the initial state $i_0$ of the
random walk is determined by sampling the source vector $\ve{b}$ with
probabilities:
\begin{equation}
  P_{i_0=i}(\nu) = \frac{b_i}{||\ve{b}||_1}\:.
  \label{eq:adjoint_source_probability}
\end{equation}
 As a result of using the adjoint system, we modify our probabilities
 and weights using the \textit{adjoint Neumann-Ulam decomposition} of
 $\ve{H}$:
\begin{equation}
  \ve{H}^{T} = \ve{P} \circ \ve{W}\:,
  \label{eq:adjoint_neumann_ulam}
\end{equation}
where now we are forming the decomposition with respect to the
transpose of $\ve{H}$. We then follow the same procedure as the direct
method for forming the probability and weight matrices in the
decomposition. Using the adjoint form, probabilities should instead be
column-scaled:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
  \label{eq:adjoint_probability}
\end{equation}
such that we expect to select a new state, $j$, from the current state
in the random walk, $i$, by sampling column-wise (or row-wise if an
adjoint probability matrix is formed). Per
Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
defined as:
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint_weight}
\end{equation}
Using the decomposition we can then define an expectation value for
the adjoint method. Given Eq~(\ref{eq:direct_permutation_weight}) as
the weight generated for a particular random walk permutation as in
Eq~(\ref{eq:mc_walk_permutation}) and our result from
Eq~(\ref{eq:inner_product_constraint}) generated by applying the
adjoint constraints, the contribution to the solution in state $j$
from a particular random walk permutation of $k$ events is then the
\textit{collision estimator}:
\begin{equation}
  X_{j}(\nu) = \sum_{m=0}^k W_{m} \delta_{i_m,j}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in
the current state the random walk starting weight will be $W_0 =
b_{i_0}$. Note here that the estimator in
Eq~(\ref{eq:adjoint_permutation_contribution}) does not have a
dependency on the source state as in
Eq~(\ref{eq:adjoint_permutation_contribution}), providing a remedy for
the situation in the direct method where we must start a random walk
in each source state for every permutation if we want to compute a
solution estimate for that state. In the adjoint method, we instead
tally in all states and those of lesser importance will not be visited
as frequently by the random walk. Finally, the expectation value using
all permutations is:
\begin{equation}
  E\{X_j\} = \sum_{\nu} P_{\nu} X_{j}(\nu)\:
  \label{eq:adjoint_expectation_value}
\end{equation}
which, if expanded in the same way as the direct method, directly
recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}
    \delta_{i_k,j} \\ &= x_{j}\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
therefore also providing an unbiased Monte Carlo estimate of the
solution. It should be noted here that
Eq~(\ref{eq:adjoint_expectation_expansion}) only computes a single
component of our desired solution vector when really what we desire is
the entire solution vector. In an adjoint Monte Carlo simulation using
this estimator, the $w_{ij}$ elements that are added into the tally
for each state are only selected if the random walk currently resides
in that state. Much like a mesh tally in a particle transport
simulation, we have $N$ simultaneous tallies for $\ve{A} \in
\mathbb{R}^{N \times N}$ that will yield the entire solution
vector. Based on their current state in the system, the random walks
will contribute tally $j$ in this group.

@@

\subsection{Collision Estimator Variance}
\label{subsec:forward_variance}
We can compute the variance of the collision estimator in the same way
as the direct estimator for each component in the solution where now:
\begin{equation}
  \sigma_j^2 = \Big(\sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2
  \delta_{i_m,j}\Big)_j - x_j^2\:.
  \label{eq:collision_variance_1}
\end{equation}
Expanding the transition probabilities yields the same variance as the
direct estimator, this time with the terms in reverse order and the
introduction of the Kronecker delta:
\begin{equation}
  \sigma_j^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N} p_{i_k,i_{k-1}}\ldots p_{i_2,i_1} p_{i_1,i_0}
  w^2_{i_k,i_{k-1}}\ldots w^2_{i_2,i_1} w^2_{i_1,i_0}
  b^2_{i_0}\delta_{i_k,j} - x_i^2\:.
  \label{eq:collision_variance_2}
\end{equation}
If we again $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$, we then have:
\begin{equation}
  \sigma^2_j = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}g_{i_{k},i_{k-1}} \ldots
  g_{i_2,i_1}g_{i_1,i_0}b_{i_0}\delta_{i_k,j}^2 - x_i^2\:,
\label{eq:collision_variance_3}
\end{equation}
which again contains the general Neumann series for $\ve{G}$ as given
by Eq~(\ref{eq:variance_neumann_series}).  Inserting $T$ back into the
variance formulation gives:
\begin{equation}
  \sigma^2_j = (\ve{T}\ve{b})_j - x_i^2\:.
  \label{eq:direct_variance_5}
\end{equation}
We can relate $\ve{G}$ to $\ve{H}$ by noting that $\ve{G}$ simply
contains an additional Hadamard product with the weight matrix. The
Hadamard product has the property that:
\begin{equation}
  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:.
  \label{eq:hadamard_inequality}
\end{equation}
Using the norm property of the Hadamard product and
Eq~(\ref{eq:neumann_ulam_decomposition}), we can define the norm of
$\ve{W}$ as:
\begin{equation}
  \frac{|\ve{H}|}{|\ve{P}|} \geq |\ve{W}|\:.
  \label{eq:weight_norm}
\end{equation}
Choosing the infinity norm of the operator as defined in
Eq~(\ref{eq:matrix_infinity_norm}), the row normalized probability
matrix will yield a norm of 1 giving the following inequality for
relating $\ve{G}$ and $\ve{H}$:
\begin{equation}
  |\ve{G}| \geq |\ve{H}|^2
  \label{eq:variance_norm_inequality}
\end{equation}
Using these relations to analyze Eq~(\ref{eq:direct_variance_5}), we
see that if $\rho(\ve{G}) > 1$, then the sum in
Eq~(\ref{eq:variance_neumann_series}) will not converge and an
infinite variance will arise as the elements of $\ve{T}$ become
infinite in Eq~(\ref{eq:direct_variance_5}). We must restrict $\ve{G}$
to alleviate this and therefore restrict $\ve{H}$ due to
Eq~(\ref{eq:variance_norm_inequality}) with $\rho(\ve{H}) < 1$ so that
our expectation values for the solution may have a finite variance.

\subsection{Expected Value Estimator}
\label{subsec:expected_value_estimator}
In addition to the collision estimator, an additional estimator is
available due to the work of Okten \citep{okten_solving_2005} that
uses the method of expected values as a means to improve the Monte
Carlo estimate. As outlined by Spanier and Gelbard
\citep{spanier_monte_1969}, the method of expected values is a
deterministic averaging of events that may potentially occur in the
Monte Carlo random walk sequence. Okten applied this principle
directly to discrete Monte Carlo by forming the \textit{expected value
  estimator} for a random walk of $k$ events:
\begin{equation}
  X_{j}(\nu) = b_j + \sum_{m=0}^k W_m h_{j,i_m}\,
  \label{eq:expected_value_estimator}
\end{equation}
where now the contribution of the iteration matrix is
deterministically averaged at step $m$ over all potentially states $j$
that may be reached from the current state $i_m$. Via Okten, the
estimator can be shown to be unbiased through a comparison to
the expected value estimator. We can first rewrite the summation in
Eq~(\ref{eq:expected_value_estimator}):
\begin{equation}
  X_{j}(\nu) = b_j + \sum_{m=0}^k \sum_{i=1}^N W_m
  \delta_{i_m,i} h_{ji}\,
  \label{eq:unbiased_eval_1}
\end{equation}
where $N$ is the number of states in the system. Immediately, we see
the collision estimator as defined by
Eq~(\ref{eq:adjoint_permutation_contribution}) and can therefore write
the expectation value as:
\begin{equation}
  E\{X_{j}\} = b_j + \sum_{i=1}^N E\{X_{i}\} h_{ji}\,
  \label{eq:unbiased_eval_2}
\end{equation}
which is equivalently is the $j^{th}$ component of
Eq~(\ref{eq:richardson_split}):
\begin{equation}
  E\{X_{j}\} = b_j + \sum_{i=1}^N x_{i} h_{ji}\,
  \label{eq:unbiased_eval_2}
\end{equation}
and is therefore and unbiased estimate. Compared to the collision
estimator, the expected value estimator provides additional
information at every step of the random walk, yielding potentially
better statistics with the same amount of transport
work. Conveniently, even if no Monte Carlo histories are computed, the
expected value estimator still deterministically computes the first
term of the Neumann Series, $\ve{H}^0\ve{b}$, whereas the collision
estimator will provide no information.

@@

@@

\subsection{Preconditioning MCSA}
\label{subsec:stochastic_preconditioning}
In most cases, at least a minimal amount of \textit{preconditioning}
of the linear system will be required in order to use the class of
stochastic methods described. Although these methods have no symmetry
requirements for convergence, they do require that the spectral radius
of the iteration matrix be less than one. Preconditioning serves as a
means of achieving this.

\subsubsection{Basic Preconditioning}
\label{subsubsec:basic_mcsa_preconditioning}
As an example, to achieve this for diagonally dominant matrices point
Jacobi left preconditioning can be used such that the preconditioning
matrix $\ve{M}$ is:
\begin{equation}
  \ve{M} = diag(\ve{A})\:.
  \label{eq:jacobi_preconditioner}
\end{equation}
With the application of this preconditioner we are instead solving the
following linear system:
\begin{equation}
  \ve{M}^{-1}\ve{A}\ve{x} = \ve{M}^{-1}\ve{b}\:.
  \label{eq:jacobi_precond_linear_problem}
\end{equation}
Next, we can apply MCSA to solve
Eq~(\ref{eq:jacobi_precond_linear_problem}): 
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k +
    \ve{M}^{-1}\ve{r}^k\:,\\ \ve{r}^{k+1/2} =
    \ve{b}-\ve{A}\ve{x}^{k+1/2}\:,\\ \ve{M}^{-1}\ve{A}\delta\ve{x}^{k+1/2}
    = \ve{M}^{-1}\ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} +
    \delta \ve{x}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
    \label{eq:jacobi_preconditioned_mcsa}
  \end{gather}
\end{subequations}
where the adjoint Monte Carlo solve now has a preconditioned operator
from which to build weights and probabilities for transport and a
preconditioned source vector to sample.

@@

\subsubsection{General Preconditioning Strategies}
\label{subsubsec:general_mcsa_preconditioning}
It is possible to use general left, right, and left/right
preconditioning with MCSA by carefully considering the underly Monte
Carlo that will be solved with the Neumann Ulam method. We consider
here the general left/right preconditioned method as the left or right
preconditioned methods can be inferred from its formulation. We
consider a left preconditioner $\ve{M_L}$ and a right preconditioner
$\ve{M_R}$. The left/right preconditioned linear problem is then:
\begin{equation}
  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{M}_R\ve{x} = \ve{M}_L^{-1}\ve{b}\:.
  \label{eq:left_right_linear_problem}
\end{equation}
To handle the right preconditioning, the system is written with a
substitution of variables:
\begin{equation}
  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{u} = \ve{M}_L^{-1}\ve{b}\:,
  \label{eq:left_right_subs_problem}
\end{equation}
with
\begin{equation}
  \ve{x} = \ve{M}_R^{-1}\ve{u}\:.
  \label{eq:left_right_recover}
\end{equation}
To apply such a method to MCSA, we solve for the substituted variable
$\ve{u}$ during the iteration sequence:
\begin{subequations}
  \begin{gather}
    \ve{u}^{k+1/2} = \ve{u}^k + \ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1/2})\:,\\ 
    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,\\ 
    \ve{u}^{k+1} = \ve{u}^{k+1/2} + \delta \ve{u}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1})\:,
    \label{eq:left_right_mcsa}
  \end{gather}
\end{subequations}
and then recover the original solution vector with
Eq~(\ref{eq:left_right_recover}). For the Monte Carlo problem, we
isolate the generation of the correction:
\begin{equation}
  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,
  \label{eq:left_right_correction}
\end{equation}
and note that the precondtioned residual of the substituted variable
is now serving as the source and the new iteration matrix is:
\begin{equation}
  \ve{H} = \ve{I} - \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\:.
  \label{eq:left_right_iteration_matrix}
\end{equation}
As we require $(i,j)$ element-wise access to the iteration matrix in
order to construct probabilities and weights for the Monte Carlo
procedure, the \textit{composite operator},
$\ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}$, must be formed via matrix-matrix
multiplication. Several possible shortcomings of this preconditioning
approach are readily observed. First, the matrix-matrix multiplication
operation for sparse, parallel distributed matrices is significantly
more expensive than a matrix-vector multiplication operation. Second,
each preconditioner must be explicitly inverted, an operation in
itself that my be expensive prohibits the use of any preconditioners
which prohibit the extraction of the inverse. Third, for many modern
preconditioning methods, this inversion may yield dense matrices,
destroying sparsity and further impeding the performance of a
matrix-matrix multiplication operation. It is also interesting to note
that the Monte Carlo problem in the general left/right preconditioned
scheme given by Eq~(\ref{eq:left_right_correction}) is not fully
left/right preconditioned (meaning that we do not recover $\ve{x}$),
but instead part of a sequence for finding the substituted variable
$\ve{u}$. We do, however, gain the benefits of this general
preconditioning by building the iteration matrix in
Eq~(\ref{eq:left_right_iteration_matrix}) from the fully
preconditioned linear operator. In addition, for MCSA to apply to
increasingly difficult problems, more advanced preconditioning
techniques that require the generation of the composite operator may
be necessary for convergence.

@@

\begin{figure}[ht!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/seq_cpu.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
      an $N \times N$ square mesh).} \textit{Both the Sequential Monte
      Carlo and MCSA solvers are used with the five point stencils and
      the adjoint Monte Carlo solver. The number of random walks was
      twice the number of discrete states in the system in order to
      ensure convergence in the Sequential Monte Carlo method.}}
  \label{fig:seq_cpu_time}
\end{figure}

\begin{figure}[ht!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/seq_iterations.pdf}
  \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
      $N \times N$ square mesh).} \textit{Both the Sequential Monte
      Carlo and MCSA solvers are used with the five point stencils and
      the adjoint Monte Carlo solver.}}
  \label{fig:seq_iterations}
\end{figure}
